Correctly Optimizing Abstract Abstract Machines

### 2013-03-29 paper/paper.tex
Changed abstract states to contexts. Tightened section on preallocation.

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, preallocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}

@@

\begin{figure}
\small
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \setof{ \inject(\compile{\mexp}^\mtcntr_\bot(\bot,\epsilon,\kmt)) \multimachstep \mstate}
                    \text{ where } \\
\inject(\mastate,\msdiff) &= \wn(\mastate,\replay(\msdiff,\bot)) \\
\wn(\mastate,\msto) \machstep \wn(\mastate',\msto') &\iff \mastate \cmachstep_\msto \mastate',\msdiff \\
\msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
\end{align*}
\\[2mm]
\begin{align*}
%% CONTINUE
\co{\kmt,\mval} &\cmachstep_\msto
\ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\mkaddr},\mval} & \cmachstep_\msto
\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}) \\
\text{ where } \maddr_f &= \alloc(\mstate) \\
               \msdiff &= \cons{\ttuple{\maddr_f}{\force(\msto,\mval)}}{\epsilon}
\\
\co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
\ap[^\mcntr_\mlab]{\mval,\mval\mkont},\epsilon \\
\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue} & \cmachstep_\msto
\mcompi{0}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse} & \cmachstep_\msto
\mcompi{1}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\mkont} & \cmachstep_\msto
\mcomp^{\mcntr'}(\msto)(\menv',\msdiff,\mkont) \\
\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
              \msdiff &= \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon}
\\
\ap{\mop,\mval,\mkont} & \cmachstep_\msto
\co{\mkont,\mval'},\epsilon \\
\text{ where } \mvalx{u} &\in \spchoice{\force(\msto,\mval)}{\force(\mval)}, \mval'\in\interpdelta(\mop,\mvalx{u})
\end{align*}
\end{gather*}
\caption{Abstract abstract machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

@@

Before we prove this setup is indeed a WEB, we need one lemma that
applying an abstract bytecode's function is equal to refining the
corresponding $\ev{}$ state:
%
\begin{lemma}[Compile/commit]
Let $\mastate,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
Let $\wn(\mastate',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
$\wn(\mastate,\replay(\msdiff',\msto)) = \wn(\mastate',\replay(\msdiff,\msto'))$.
\end{lemma}
The proof is by induction on $\mexp$.

@@

\subsection{Imperative, preallocated data structures}

@@

{\small
\begin{align*}
\System &= (\widehat{\State} \to {\Timestamp}^*) \times \wp(\widehat{\State}) \times \Store \times \Timestamp \\
{\mathcal F} &: \System \to \System \\
{\mathcal F}(S,F, \msto,t) &= (S',F',\msto', t') \\
\text{ where }
I &= \setof{(\mastate',\msdiff) \mid
       \mastate \in F,
       \mastate \cmachstep_{\msto^*} \mastate',\msdiff} \\
\msto^* &= \lambda \maddr.\lookup(\msto(\maddr),t) \\
\ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
t' &= \left\{\begin{array}{ll} t+1 & \text{ if } \changep \\
              t   & \text{ otherwise}
             \end{array}\right. \\
F' &= \setof{\mastate \mid (\mastate,\_) \in I, \changep \vee S(\mastate) \neq \cons{t}{\_}} \\
S' &= \lambda \mastate. \left\{\begin{array}{ll}
                               \cons{t'}{S(\mastate)} & \text{ if } \mastate \in F' \\
                               S(\mastate) & \text{ otherwise}
                             \end{array}\right.
\end{align*}}

@@

Internally, the algorithm at this stage uses hash tables to model the store.
%
This is because stores used to be distributed to all states, which
required a compact, dynamic representation.
%
But, such a dynamic structure isn't necessary when we know the
structure of the store in advance: we know all possible entries, and
we know its maximum size.

In a monovariant analysis, the domain of the store is
exactly the set of expressions in the program.
%
If we label each expression with a unique natural, the analysis can
index directly into the store without a hash or a collision.
%
Even for polyvariant analyses, it is possible to compute the maximum
number of addresses and similarly pre-allocate either the spine of the
store or (if memory is no concern) the entire store.

>>>>

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, pre-allocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}

@@

\begin{figure}
\small
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \setof{ \inject(\compile{\mexp}^\mtcntr_\bot(\bot,\epsilon,\kmt)) \multimachstep \mstate}
                    \text{ where } \\
\inject(c,\msdiff) &= \wn(c,\replay(\msdiff,\bot)) \\
\wn(c,\msto) \machstep \wn(c',\msto') &\iff c \cmachstep_\msto c',\msdiff \\
\msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
\end{align*}
\\[2mm]
\begin{align*}
%% CONTINUE
\co{\kmt,\mval} &\cmachstep_\msto
\ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\mkaddr},\mval} & \cmachstep_\msto
\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}) \\
\text{ where } \maddr_f &= \alloc(\mstate) \\
               \msdiff &= \cons{\ttuple{\maddr_f}{\force(\msto,\mval)}}{\epsilon}
\\
\co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
\ap[^\mcntr_\mlab]{\mval,\mval\mkont},\epsilon \\
\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue} & \cmachstep_\msto
\mcompi{0}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse} & \cmachstep_\msto
\mcompi{1}^\mcntr(\msto)(\menv,\epsilon,\mkont)
\text{ if }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\mkont} & \cmachstep_\msto
\mcomp^{\mcntr'}(\msto)(\menv',\msdiff,\mkont) \\
\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
              \msdiff &= \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon}
\\
\ap{\mop,\mval,\mkont} & \cmachstep_\msto
\co{\mkont,\mval'},\epsilon \\
\text{ where } \mvalx{u} &\in \spchoice{\force(\msto,\mval)}{\force(\mval)}, \mval'\in\interpdelta(\mop,\mvalx{u})
\end{align*}
\end{gather*}
\caption{Abstract abstract machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

@@

Before we prove this setup is indeed a WEB, we need one lemma that
applying an abstract bytecode's function is equal to refining the
corresponding $\ev{}$ state:
%
\begin{lemma}[Compile/commit]
Let $c,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
Let $\wn(c',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
$\wn(c,\replay(\msdiff',\msto)) = \wn(c',\replay(\msdiff,\msto'))$.
\end{lemma}
The proof is by induction on $\mexp$.

@@

\subsection{Imperative, pre-allocated data structures}

@@

{\small
\begin{align*}
\System &= (\widehat{\State} \to {\Timestamp}^*) \times \wp(\widehat{\State}) \times \Store \times \Timestamp \\
{\mathcal F} &: \System \to \System \\
{\mathcal F}(S,F, \msto,t) &= (S',F',\msto', t') \\
\text{ where }
I &= \setof{(c',\msdiff) \mid
       c \in F,
       c \cmachstep_{\msto^*} c',\msdiff} \\
\msto^* &= \lambda \maddr.\lookup(\msto(\maddr),t) \\
\ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
t' &= \left\{\begin{array}{ll} t+1 & \text{ if } \changep \\
              t   & \text{ otherwise}
             \end{array}\right. \\
F' &= \setof{c \mid (c,\_) \in I, \changep \vee S(c) \neq \cons{t}{\_}} \\
S' &= \lambda c. \left\{\begin{array}{ll}
                               \cons{t'}{S(c)} & \text{ if } c \in F' \\
                               S(c) & \text{ otherwise}
                             \end{array}\right.
\end{align*}}

@@

Internally, the algorithm at this stage uses hash tables to model the store to allow arbitrary address representations.
%
But, such a dynamic structure isn't necessary when we know the
structure of the store in advance: we know all possible entries, and
we know its maximum size.
%
In a monovariant allocation strategy, the domain of the store is
exactly the set of expressions in the program.
%
If we label each expression with a unique natural, the analysis can
index directly into the store without a hash or a collision.
%
Even for polyvariant analyses, it is possible to compute the maximum
number of addresses and similarly pre-allocate either the spine of the
store or (if memory is no concern) the entire store.
